--- 
title: "Tidy Tuesday Cetaceans"
subtitle: "Kate Straneva"
output:
  html_document:
    df_print: paged
---

# Packages

```{r packages}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(tibble)
library(lubridate)
library(RColorBrewer)
library(formatR)
```

# Data

```{r load_data}
cetaceans <- read_csv("allCetaceanData.csv")
dim(cetaceans)
head(cetaceans)
```



### Data Exploration & Analysis ### 

```{r include = TRUE}

eightiesfish <- cetaceans%>% 
  filter(birthYear >= 1980 ) %>%
  filter(birthYear <= 1989) %>%
  filter(region == "US")


# Number of cars in each class:

## plto the birthyear
ggplot(eightiesfish, aes(birthYear)) + geom_histogram(binwidth = 1) + ggtitle(label = "Distribution of births by Year", subtitle = "Cetaceans births in 1980s") + theme_bw()
 
#plotSex
ggplot(eightiesfish, aes(sex)) + geom_bar() + ggtitle(label = "Cetacean Gender") + theme_classic()

```

```{r}

## Find unique species values 

unique(cetaceans$species)

## Find the number of unique species 
length(unique(cetaceans$species))

```

# There are 37 unique species that are represented in the dataframe 

# Examine all the unique ways in which species were Acquired 

```{r}
unique(cetaceans$acquisition)


```
```{r}
ggplot(cetaceans, aes(acquisition)) + geom_bar() + ggtitle(label = "Acquistion Method") + theme_classic()

```

The majority of acquisitons occured through caputre, followed by being born in captivity. 

## Plot the types of species and there relative count 

```{r }



ggplot(cetaceans, aes(species)) + geom_bar() + coord_flip() + ggtitle(label = "Distribution of Species") 

```

The most common species represented in the data is the bottlenose dolphin. 


# Gender Breakdown of Species acqured 

```{r}


ggplot(cetaceans, aes(sex)) + geom_bar() + coord_flip() + ggtitle(label = "Distribution of Genders") + theme_minimal() 


```
There were almost 300 more females than males represented in the dataset. 


# More advanced Visualizations 




## 1. Visualization comparing the average age of males versus females faceted by species. 

To complete this visualization, I first need to calculate the age of each animal that is still alive. 
I will filter the data to only include animals marked with the status "Alive." 
I will then caculate the age by subtracting 2018 from the year they were born 
(2018 because this dataset was created in 2018). I will only include the top 
four species represented in the data. I will group by species and caculate the 
average age for each species. I will use geom_boxplot to show the median age 
of each species. 


```{r}

alive <- cetaceans%>%
  filter(status == "Alive") %>% 
  filter(sex == "F" | sex == "M") %>% 
  filter(species %in% c("Beluga", "Bottlenose", "Killer Whale; Orca", "Pilot, Short-fin")) %>%
  mutate(age = 2018 - birthYear)%>% 
  select(species, age, sex, acquisition, region) %>% 
  group_by(species) %>% 
  mutate(Average_Age = mean(age)) 


##mutate(Average_Age = format(Average_Age,'%Y'))

head(alive)


ggplot(alive, aes(x = sex, y = age, fill = sex)) + geom_boxplot(outlier.colour="red", outlier.shape=8,
                outlier.size=4)   + ggtitle(label = "Average age by Gender for top Species")  + facet_grid(~species) +  scale_fill_brewer(palette="Set3") +  theme_minimal() 



```
 In this dataset the medianage for male killer whales and  male pilot whales is 
 higher than females.The median age of male and femalebottlenose dolphins
 are very similar. In this dataset, the beluga whale females had a older median 
 age than males. I chose to highlight the outliers using a red asterisk.






```